---
title: "Reproducible Research: Peer Assessment 1"
output: 
html_document:
keep_md: true
---

### Introduction:
This MD file will include the code and results to complete Coursera's **Reproducible Research** first course assignment. The code assumes that the ggplot library is installed, and that the ziped file of the data is located in the working directory.


## Loading and preprocessing the data

The data for this assignment was downloaded from the course web site:

* Dataset: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) [52K]

The variables included in this dataset are:

* **steps**: Number of steps taking in a 5-minute interval (missing values are coded as `NA`)

* **date**: The date on which the measurement was taken in YYYY-MM-DD format

* **interval**: Identifier for the 5-minute interval in which the measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

Loading the data and the ggplot2 package:

```{r}
unzip(zipfile = "activity.zip")
data <- read.csv("activity.csv")
library(ggplot2)
```

## What is mean total number of steps taken per day?

```{r}
total.steps <- tapply(data$steps, data$date, FUN=sum, na.rm=TRUE)
qplot(total.steps, binwidth=1000, xlab="total number of steps taken each day")
mean(total.steps, na.rm=TRUE)
median(total.steps, na.rm=TRUE)
```

## What is the average daily activity pattern?

```{r}
averages <- aggregate(x=list(steps=data$steps), by=list(interval=data$interval),
                      FUN=mean, na.rm=TRUE)
ggplot(data=averages, aes(x=interval, y=steps)) +
    geom_line() +
    xlab("5-minute interval") +
    ylab("average number of steps taken")
```

On average across all the days in the dataset, the 5-minute interval that contains the maximum number of steps is:

```{r}
averages[which.max(averages$steps),]
```

## Imputing missing values

Since the raw data contains many days and intervals where there are missing values (`NA`), we remove these missing values to prevent skewed results in our calculations:

```{r}
missing <- is.na(data$steps)
table(missing)
```

Replacing missing values with the mean value for that 5-minute interval:

```{r}
fill.value <- function(steps, interval) {
    filled <- NA
    if (!is.na(steps))
        filled <- c(steps)
    else
        filled <- (averages[averages$interval==interval, "steps"])
    return(filled)
}
filled.data <- data
filled.data$steps <- mapply(fill.value, filled.data$steps, filled.data$interval)
```

Creating a histogram that shows the total number of steps taken each day, as well as the mean and median total number of steps.

```{r}
total.steps <- tapply(filled.data$steps, filled.data$date, FUN=sum)
qplot(total.steps, binwidth=1000, xlab="total number of steps taken each day")
mean(total.steps)
median(total.steps)
```

The mean and median values are higher when using the imputed data set. This is because the original data set contained several days where all intervals had 0 steps recorded. In the imputed data set, this is not the case and the calculated value for the mean and median is higher and more accurate. 

## Are there differences in activity patterns between weekdays and weekends?

using the weekdays function to asign a day of the week for each measurement in the imputed data set:

```{r}
weekday.or.weekend <- function(date) {
    day <- weekdays(date)
    if (day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
        return("weekday")
    else if (day %in% c("Saturday", "Sunday"))
        return("weekend")
    else
        stop("invalid date")
}
filled.data$date <- as.Date(filled.data$date)
filled.data$day <- sapply(filled.data$date, FUN=weekday.or.weekend)
```

Creating a panel plot for the average number of steps taken on weekdays and weekends:
```{r}
averages <- aggregate(steps ~ interval + day, data=filled.data, mean)
ggplot(averages, aes(interval, steps)) + geom_line() + facet_grid(day ~ .) +
    xlab("5-minute interval") + ylab("Number of steps")
```
